Galaxy Imaging with Generative Models: Insights from a Two-Models Framework

### 2025-03-13 MNRAS/galaxy-generative-models-overleaf-submit-2nd/article_MNRAS.tex
start sec 3

{\color{red}The central challenge in generation lies in estimating the \textit{score}, the key ingredient of the generation process.} Two main approaches have been proposed: (1) deriving a model for $p_t$ that incorporates regularity patterns and correlations to learn the \textit{score} via \textit{score matching} \citep{hyvarinen2005a}, or (2) employing a \textit{denoising} neural network, as introduced by \cite{Bengio2013}, further developed in \cite{Sohl-Dickstein2015,Ho2020}, and recently examined from a mathematical perspective in \cite{kadkhodaie2024generalization}. The first approach has been utilized in works such as \cite{Guth2022b,Lempereur2024} with Gibbs energy parameterization. The second has found applications in astrophysics, including galaxy image generation \citep{smith2021}, generation of 21~cm luminosity temperature maps \citep{Zhao2023}, super-resolution of large-scale cosmic structures \citep{Schanz2023}, and Bayesian posterior sampling for weak lensing mass-mapping problems \citep{Remy2023}.

@@

{\color{red}
\subsection{General comments}
%

@@

Returning to the original GAN problem, the question \textit{"Does $p_G = p_{\text{data}}$?"} remains an open challenge. This is why numerical experiments are still a valuable approach. We have selected three models, each representing one of the three types described in this section, to assess the reliability of model optimization. Regarding the number of model parameters, we maintain them in the range of $O(10)$ million, which is significantly lower than the scale used in large diffusion models mentioned in the introduction. This choice is relevant for two reasons: first, it aligns with the constraints faced by practitioners with limited computational resources; second, it is well-suited to the size of the images ($d \times d$) and the dataset ($N$) used in our experiments.

Anticipating one of the results, we observe that, as a general rule of thumb, the total number of available training pixels ($N d^2$) should be at least on the order of $O(10^2)$ times the total number of model parameters to ensure effective learning.
}

@@

%
\section{Experiment}
\label{sec-experiment}
%
In the previous section, we briefly outlined different architectures of generative models. Here, we describe the process of generating a sample $\bm{x}$, which typically\footnote{This is the usual choice, but alternatives such as uniform distributions can also be considered; see \cite{Brock2019} for a discussion in the context of GANs.} begins with sampling from a centered isotropic Gaussian distribution $\bm{z}\sim \mathcal{N}(\bm{0}, \bm{1})$ (i.e., white noise) of dimension $d_\ell$. For VAE/GAN-based models, $d_\ell \ll d$, while $d = d_\ell$ for flow-based and diffusion-based models.

@@

\subsection{The dataset}
\label{sec-Dataset}
%
For training, \cite{kadkhodaie2024generalization} employed widely used machine learning datasets, including \texttt{CelebA} \citep{Liu2015}, \texttt{LSUN} Bedroom \citep{Yu2015}, and \texttt{CelebA HQ} \citep{KarrasALL18}. All images were grayscale (single channel) with intensity values in the range $[0,1]$. The \texttt{LSUN} images were downsampled to a resolution of $80 \times 80$ pixels, while \texttt{CelebA} images were resized to $32 \times 32$ and \texttt{CelebA HQ} images to $40 \times 40$ pixels. The transition from memorization to generalization was observed with a training dataset consisting of $10^5$ images.

In the present study, we focus on galaxy image generation. Given the requirement to train models with at least $10^5$ images, we used the Sloan Digital Sky Survey (SDSS) dataset described in \citep{smith2021}\footnote{Raw data available at \url{https://github.com/Smith42/astroddpm}.} with modifications to preprocessing. Specifically, all images were cropped around target galaxy coordinates to $64 \times 64$ pixels, and a single grayscale channel was created from the $(z, g, r)$ channels using the \texttt{scipy} \citep{2020SciPy-NMeth} library\footnote{The \texttt{make\_lupton\_rgb} function was applied with $Q=8$ and \texttt{stretch}=0.2 settings.}. 
%
\subsection{The models}
\label{sec-Exp-Models}
%
We selected three architectures, as described in Section~\ref{sec-generative-models}: a GAN, a flow-based model, and a diffusion-based model. We have used implementations of architectures written in \texttt{PyTorch} \citep{PyTorch2019}.

>>>>

{\color{red}The central challenge in generation lies in estimating the \textit{score}, the key ingredient of the generation process.} Two main approaches have been proposed: (1) deriving a model for $p_t$ that incorporates regularity patterns and correlations to learn the \textit{score} via \textit{score matching} \citep{hyvarinen2005a}, or (2) employing a \textit{denoising} neural network, as introduced by \cite{Bengio2013}, further developed in \cite{Sohl-Dickstein2015,Ho2020,song2020score}, and recently examined from a mathematical perspective in \cite{kadkhodaie2024generalization}. The first approach has been utilized in works such as \cite{Guth2022b,Lempereur2024} with Gibbs energy parameterization. The second has found applications in astrophysics, including galaxy image generation \citep{smith2021}, generation of 21~cm luminosity temperature maps \citep{Zhao2023}, super-resolution of large-scale cosmic structures \citep{Schanz2023}, and Bayesian posterior sampling for weak lensing mass-mapping problems \citep{Remy2023}.

@@

{\color{red}
\subsection{Some comments}
%

@@

Returning to the original GAN problem, the question \textit{"Does $p_G = p_{\text{data}}$?"} remains an open challenge. This is why numerical experiments are still a valuable approach. We have selected three models, each representing one of the three types described in this section, to assess the reliability of model optimization. This choice may be questioned but we think that it serves more as guide lines if one would like to use other models. 
Regarding the number of model parameters, we maintain them in the range of $O(10)$ million, which is significantly lower than the scale used in large diffusion models mentioned in the introduction. This choice is relevant for two reasons: first, it aligns with the constraints faced by practitioners with limited computational resources; second, it is well-suited to the sizes of the images ($d \times d$) and the dataset ($N$) used in our experiments.  
Anticipating one of the results, we observe that, as a general rule of thumb, the ratio of the total number of available training pixels to the total number of model parameters should be at least on the order of $O(10^2)$ to ensure effective learning.
}

@@

%
\section{Experiment}
\label{sec-experiment}
%
In the previous section, we briefly outlined different architectures of generative models. Here, we describe the process of generating a sample $\bm{x}$, which typically\footnote{This is the usual choice, but alternatives such as uniform distributions can also be considered; see \cite{Brock2019} for a discussion in the context of GANs.} begins with sampling from a centered isotropic Gaussian distribution $\bm{z} \sim \mathcal{N}(\bm{0}, \bm{1})$ (i.e., white noise) of dimension $d_\ell$. {\color{red} Notably, unlike GANs and VAEs, where latent space compression occurs ($d_\ell \ll d$), diffusion and flow-based models preserve the data dimensionality ($d = d_\ell$).}

@@

\subsection{The dataset}
\label{sec-Dataset}
%
%13-mars-25
%For training, \cite{kadkhodaie2024generalization} employed widely used machine learning datasets, including \texttt{CelebA} \citep{Liu2015}, \texttt{LSUN} Bedroom \citep{Yu2015}, and \texttt{CelebA HQ} \citep{KarrasALL18}. All images were grayscale (single channel) with intensity values in the range $[0,1]$. The \texttt{LSUN} images were downsampled to a resolution of $80 \times 80$ pixels, while \texttt{CelebA} images were resized to $32 \times 32$ and \texttt{CelebA HQ} images to $40 \times 40$ pixels. The transition from memorization to generalization was observed with a training dataset consisting of $10^5$ images.
%
%In the present study, we focus on galaxy image generation. Given the requirement to train models with at least $10^5$ images, we used the Sloan Digital Sky Survey (SDSS) dataset described in \citep{smith2021}\footnote{Raw data available at \url{https://github.com/Smith42/astroddpm}.} with modifications to preprocessing. Specifically, all images were cropped around target galaxy coordinates to $64 \times 64$ pixels, and a single grayscale channel was created from the $(z, g, r)$ channels using the \texttt{scipy} \citep{2020SciPy-NMeth} library\footnote{The \texttt{make\_lupton\_rgb} function was applied with $Q=8$ and \texttt{stretch}=0.2 settings.}.

{\color{red}
In the present study, we focus on galaxy image generation, having as underlying guide the experiment performed by \cite{kadkhodaie2024generalization} who  employed widely used machine learning datasets. All images were grayscale (single channel) with intensity values in the range $[0,1]$ and resolution of $80 \times 80$ pixels. The transition from memorization to generalization was observed with a training dataset consisting of $10^5$ images.

Given the requirement to train models with at least $10^5$ images, we used the Sloan Digital Sky Survey (SDSS) dataset described in \citep{smith2021}\footnote{Raw data available at \url{https://github.com/Smith42/astroddpm}.} with modifications to preprocessing. Specifically, all images were cropped around target galaxy coordinates to $64 \times 64$ pixels, and a single grayscale channel was created from the $(z, g, r)$ channels using the \texttt{scipy} \citep{2020SciPy-NMeth} library\footnote{The \texttt{make\_lupton\_rgb} function was applied with $Q=8$ and \texttt{stretch}=0.2 settings.}. 
}
%
\subsection{The models}
\label{sec-Exp-Models}
%
We selected three architectures, as described in Section~\ref{sec-generative-models}: a GAN, a flow-based model, and a diffusion-based model. We have used implementations of architectures written in \texttt{PyTorch} \citep{PyTorch2019}.
